Defining properties regulating pioneer-factor binding and function

### 2023-01-18 manuscript/manuscript_sections/_02_Introduction.qmd
add introduction draft

@@

outline:

-   Introduction to pioneer factors: what are they and why do they matter

-   articulate the knowledge gap and questions: how do PFs select their binding sites?
    What controls context-dependent pioneer factor binding and activity?

-   Introduce Zld and Grh.

-   Summary paragraph.

>>>>

---
editor: 
  markdown: 
    wrap: 72
---

@@

Transcription factors activate transcription by binding to specific DNA
motifs and recruiting transcriptional machinery. The packaging of DNA
into chromatin obstructs TFs from accessing their target motifs. Thus,
many TFs can only bind to motifs that fall within nucleosome-free
regions (NFRs). One mechanism for how NFRs are established is through
the action of pioneer transcription factors (PFs). PFs are defined by
their ability to recognize target motifs on nucleosomes. PFs bind to
nucleosomal targets, promote chromatin opening, and create NFRs that
allow additional TFs to bind and activate transcription. Through this
activity, PFs drive cell-type specific gene expression programs that
determine cell fate. Due to these properties, PFs are essential players
in development, induced reprogramming, and disease.

Despite the importance of PFs to these diverse biological processes, how
PFs select their binding sites in different biological contexts remains
incompletely understood. Despite their ability to target inaccessible,
nucleosomal binding sites, PFs do not bind to all genomic instances of
their target motifs. Many PFs have also been shown to have cell-type
specific patterns of binding and activity. How PF binding is restricted
in different tissues or cellular contexts remains unclear. Many PFs bind
primarily to naïve chromatin that is inaccessible, inactive, and lacks
most histone modifications. Cell-type specific binding is often
associated with cell-type specific chromatin features, however many of
these observations have been correlative, and this has not been
extensively tested functionally. In some cases, inhibition of various
chromatin modifiers can enhance the ability of PFs to promote cellular
reprogramming, but the molecular basis of how this relates to PF binding
is unclear. Thus, a more systematic exploration of how PF binding and
activity is regulated is essential to understanding how these factors
function in development and disease.

Two well studied PFs that display context-dependent activity and Zelda
(Zld) and Grainy head (Grh). Zld is essential for driving early
embryonic development in Drosophila. Immediately following
fertilization, the zygotic genome is transcriptionally silent and
development is controlled by maternally deposited mRNAs and proteins.
Maternal products are gradually degraded as the zygotic genome is
activated during a process known as the maternal-to-zygotic transition
(MZT). In Drosophila, the pioneering activity Zld is essential for
driving the MZT. Zld was the first discovered activator of the zygotic
genome, and work on Zld paved the way for the identification of
analogous proteins in other species. Our lab recently identified an
additional function of Zld in promoting neural stem cell (NSC) fate in
the developing larval brain. The majority of Zld binding sites are
unique to either the embryo or NSCs, with only a subset of regions being
bound in both tissues. How Zld is directed to distinct binding sites in
the embryo and NSCs in unknown.

Grh is a master regulator of epithelial cell fate that is conserved
among metazoans. In Drosophila, Grh acts as a pioneer factor in late
embryos and larval tissues. Unlike Zld, Grh binding is similar across
different tissues and developmental stages. Despite this stable binding,
Grh is dispensible for chromatin accessibility in the early embryo, and
only displays pioneering activity in late embryos or larval tissues.
This intriguing combination of stable binding and cell-type specific
activity raises questions about how pioneering by Grh is regulated
during development. Additionally, overexpression of Grh is a hallmark of
many epithelial cancers. The role of Grh in cancer is not well
understood, as Grh can either inhibit or promote tumorigenesis,
depending on the context.

The essential roles of Zld and Grh in diverse biological processes and
the well documented context-dependent binding and activity of these
factors makes them excellent models to explore how PF binding and
activity are regulated. Because many previous studies have relied on
loss-of-function experiments, it has been difficult to unravel causal
relationships between chromatin structure and PF activity. We therefore
expressed Zld and Grh in Drosophila S2 cells, where neither factor is
endogenously expressed. We demonstrate that when expressed ectopically,
Zld and Grh bind to closed chromatin, establish NFRs, and promote
transcriptional activation. We show that an additional TF, Twist (Twi),
can act as a pioneer factor. Pioneering by Zld, Grh, and Twi depends on
a combination of local chromatin structure, sequence context, and PF
concentration and requires protein regions outside the DNA-binding
domain. We propose that pioneering activity is not a binary feature
intrinsic to a protein, but occurs on a spectrum and can be regulated by
a variety of protein-intrinsic and cell-type specific features.

<!--# Should I discuss what is known about DBDs, structural features, and how PFs engage chromatin? -->
